Experiences with Planning for Natural Language Generation

### 2009-06-01 trunk/Papers/ci-spark/conclusion.tex
fixed conclusion

In this paper, we introduced two novel planning domains arising from
problems in natural language generation: the sentence generation domain and
the Generating Instructions in Virtual Environments (GIVE) challenge
domain. We also reported the results of a set of experiments designed to
evaluate the suitability of off-the-shelf planners (particularly FF and
SGPLAN) in these domains.

Our results were mixed. While the planners we tested do an impressive job
of controlling the complexity of search in many domains, they also suffer
from practical problems that limit their performance in unexpected ways: in
both of our NLG domains the grounding step performed by FF and SGPLAN
dominates the time it takes to perform the search itself. As future work,
we hope to extend our analysis to other planners that do not perform the
initial grounding step, planners such as SHOP2
\citep{DBLP:journals/jair/NauAIKMWY03} that have traditionally performed
well on a wide range of domains, and new planners introduced in the latest
edition of the IPC.\footnote{See
 \url{http://ipc.informatik.uni-freiburg.de/} for details of the 2008
 International Planning Competition (deterministic track).}

The NLG community's recent interest in planning presents a valuable
opportunity for planning researchers. Domains like GIVE highlight certain
challenges, such as plan execution monitoring and plan presentation (i.e.,
summarisation and elaboration), but also offer a platform on which such
technologies can be evaluated in experiments with human users. Although we
have focused on classical planning problems in this work, research related
to reasoning under uncertainty, resource management, and planning with
knowledge and sensing, can also be investigated in these settings. As such,
we believe our domains would provide suitable challenges for planners
entered in future editions of the IPC. Provided some of the challenges we
have identified can be addressed, NLG-inspired problem domains offer a
constructive platform for the planning community to showcase their
techniques to a wider audience---and to improve the quality of their tools
for real-world planning tasks.

\todo{planning claims general applicability; it's clear that IPC
  benchmarks focus areas of improvement; but it's not so clear to what
  extent this limits usefulness for non-IPC application areas; this
  paper evaluates this (for the first time??)}

@@

This work arose in the context of the Planning and Language Interest Group
at the University of Edinburgh. The authors would like to thank all members
of this group, especially Hector Geffner and Mark Steedman, for interesting
discussions. This work was partially supported by the DFG Research
Fellowship ``CRISP: Efficient integrated realization and microplanning''
and by the European Commission through the PACO-PLUS project
(FP6-2004-IST-4-27657).

>>>>

In this paper, we have investigated the usefulness of current planning
technology to natural language generation, an application area with a
long tradition that has recently experienced renewed interest from NLG
researchers. In particular, we have evaluated the performance of
several off-the-shelf planners on a series of planning domains that
arise in the context of sentence generation and situated instruction
generation.

Our results were mixed. While some of the planners we tested---in
particular, FF and SGPLAN, in different domains---did an impressive
job of controlling the complexity of the search, we also found that
all planners we tested spent too much time on preprocessing to be
useful. In the sentence generation domain, FF spent 90\% of its
runtime on computing the ground instances of the planning operators;
in the instruction-giving domain, which is very similar to Gridworld,
a similar effect happened once we added more predicates. As things
stand, we found that this overly long preprocessing time makes current
planners inapplicable to NLG application for any but the smallest
instances. Users who come to planning from the outside, such as NLG
researchers, take planners as a black box. This means that search
efficiency alone is not helpful when other modules of the planner are
slow. From this perspective, we propose that the planning community
might spend some attention on optimizing the preprocessing module with
similar vigor as the search itself. In particular, we propose to
investigate planning algorithms that do not have to rely on grounding
out all operators before starting the search.

NLG and planning have a long history in common. The recent surge in
NLG-as-planning research presents valuable opportunities for both
sides. Clearly, NLG researchers who apply planning technology will
benefit directly from any improvements in planner effiency. But
conversely, NLG may be a worthwhile application area for planning
researchers to keep in mind. Domains like GIVE highlight certain
challenges, such as plan execution monitoring and plan presentation
(i.e., summarisation and elaboration), but also offer a platform on
which such technologies can be evaluated in experiments with human
users. Further, although we have focused on classical planning
problems in this work, research related to reasoning under
uncertainty, resource management, and planning with knowledge and
sensing, can also be investigated in these settings. As such, we
believe our domains would provide interesting challenges for planners
entered in future editions of the IPC.

@@

This work arose in the context of the Planning and Language Interest
Group at the University of Edinburgh. The authors would like to thank
all members of this group, especially Hector Geffner and Mark
Steedman, for interesting discussions. We also thank our reviewers for
their insightful and challenging comments.  This work was supported by
the DFG Research Fellowship ``CRISP: Efficient integrated realization
and microplanning'', the DFG Cluster of Excellence ``Multimodal
Computing and Interaction'', and by the European Commission through
the PACO-PLUS project (FP6-2004-IST-4-27657).

### 2009-06-01 trunk/Papers/ci-spark/discussion.tex
fixed conclusion

This means that while current planners deal well with the search
problems we have looked at, they are still unusable for practical NLG
applications because of the time they spend on preprocessing. In other
words, generation-as-planning research is still not in a much better
position than it was in the 1980s. While we are aware that the time
that a planner invests in preprocessing can pay off during search, we
still suggest that the inability of current planners to scale to
larger domains still limits their usefulness for applications beyond
NLG as well. Furthermore, we feel that preprocessing gets less
research attention than it deserves: If it is scientifically trivial,
we challenge the planning community to come up with implementations
that only ground operators by need; otherwise, we look forward to
publications on this topic. To support this, we offer our planning
domains as benchmarks for future competitions. They are available at
\todo{upload them somewhere}

>>>>

Of course, it would be premature to judge the usefulness of current
planners for NLG as a whole based on just two NLG
domains. Nevertheless, we believe that the structure of our planning
problems, which are dominated by a large number of operators and
individuals, is typical for NLG-related planning problems as a
whole. This suggests strongly that while current planners deal well
with the search problems we have looked at, they are still unusable
for practical NLG applications because of the time they spend on
preprocessing. In other words, generation-as-planning research is
still not in a much better position than it was in the 1980s. While we
are aware that the time that a planner invests in preprocessing can
pay off during search, we still suggest that the inability of current
planners to scale to larger domains still limits their usefulness for
applications beyond NLG as well. Furthermore, we feel that
preprocessing gets less research attention than it deserves: If it is
scientifically trivial, we challenge the planning community to come up
with implementations that only ground operators by need; otherwise, we
look forward to publications on this topic. To support this, we offer
our planning domains as benchmarks for future competitions. They are
available at \todo{upload them somewhere}

### 2009-06-01 trunk/Papers/ci-spark/introduction.tex
fixed conclusion

Treating generation as planning has a long history in NLG, ranging from the
initial attempts of the field to utilise early planning approaches
\citep{perrault80,appelt:planning,hovy88,young94dpocl}, to the recent surge of
research \citep{Steedman-Petrick:07,KolSto07,benotti08b} seeking to capitalise
on the improvements modern planners offer in terms of efficiency and
expressiveness. This paper attempts to assess the usefulness of current planning
techniques to NLG by investigating some representative generation problems, and
evaluating whether automated planning has advanced to the point that it can
provide solutions to such NLG applications---applications that are not currently
being investigated by mainstream planning research.

The focus of this paper is twofold. First, we present two generation problems
that have recently been cast as planning problems: the sentence generation task
and the GIVE task. In the sentence generation task, we concentrate on generating
a single sentence that expresses a given meaning. In this case, a plan encodes
the necessary sentence with the actions in the plan corresponding to the
utterance of individual words \citep{KolSto07}. In the GIVE domain (``Generating
Instructions in Virtual Environments''), we describe a new shared task that was
recently posed as a challenge for the NLG community
\citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part of a larger NLG
system for generating natural-language instructions that guide a human user in
performing a given task in a virtual environment.

Second, we discuss our experiences using a set of off-the-shelf planners in
these two generation domains. Among the planners we test, we explore the
efficiency of FF \citep{HoffmannNebel01}---a planner that has arguably had the
greatest impact on recent approaches to deterministic planning---and its many
descendants, such as SGPLAN \citep{hsu06:_new_featur_in_sgplan_for}. All of the
planners we test are freely available, support an expressive subset of the
Planning Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
and have been successful on both standard planning benchmarks and the problems
of the International Planning Competition (IPC).\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about recent and past
  editions of the IPC. A good overview of the deterministic track of the 2004
  instance of the IPC can be found in \citep{Hoffmann-Edelkamp:05}.}
Using these planners---and an ad-hoc Java implementation of GraphPlan
\citep{Blum1997} which serves as a baseline for certain experiments---we perform
a series of tests on a range of problem instances in our NLG domains.

Overall, our findings are mixed. On the one hand, we demonstrate that certain
planners can readily handle the \emph{search} problems that arise in our testing
domains on realistic inputs, which is promising given the challenging nature of
these tasks (e.g., the sentence generation task is NP-complete \citep{KolStr02})
On the other hand, these same planners spend tremendous amounts of time on
\emph{preprocessing} techniques that analyse the problem domain in support of
the search. (For instance, FF spends 90\% of its runtime in the sentence
generation domain on grounding out actions and literals.) As a consequence, the
off-the-shelf planners we investigated are often too slow to be useful in real
NLG applications.  Nevertheless, we offer our NLG domains and experiences as
both challenges and lessons for the planning community, in the hope that the
issues we raise might gain more attention in the future.

>>>>

Treating generation as planning has a long history in NLG, ranging
from the initial attempts of the field to utilise early planning
approaches \citep{perrault80,appelt:planning,hovy88,young94dpocl}, to
a recent surge of research
\citep{Steedman-Petrick:07,KolSto07,brenner08:_contin_multiag_plann_approac_to_situat_dialog,benotti08b}
seeking to capitalise on the improvements modern planners offer in
terms of efficiency and expressiveness. This paper attempts to assess
the usefulness of current planning techniques to NLG by investigating
some representative generation problems, and evaluating whether
automated planning has advanced to the point that it can provide
solutions to such NLG applications---applications that are not
currently being investigated by mainstream planning research.

To answer this question, we proceed in two steps. First, we present
two generation problems that have recently been cast as planning
problems: the sentence generation task and the GIVE task. In the
sentence generation task, we concentrate on generating a single
sentence that expresses a given meaning. In this case, a plan encodes
the necessary sentence with the actions in the plan corresponding to
the utterance of individual words \citep{KolSto07}. In the GIVE domain
(``Generating Instructions in Virtual Environments''), we describe a
new shared task that was recently posed as a challenge for the NLG
community \citep{ByrKolStrCasDalMooObe09}. GIVE uses planning as part
of a larger NLG system for generating natural-language instructions
that guide a human user in performing a given task in a virtual
environment.

Second, we evaluate the performance of current off-the-shelf planners
in the planning domains into which these two generation problems
translate. Among the planners we test, we explore the efficiency of FF
\citep{HoffmannNebel01}---a planner that has arguably had the greatest
impact on recent approaches to deterministic planning---and some of its
descendants, such as SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}. All of the planners we test
are freely available, support an expressive subset of the Planning
Domain Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}),
and have been successful on both standard planning benchmarks and the
problems of the International Planning Competition (IPC).\footnote{See
  \url{http://ipc.icaps-conference.org/} for information about recent
  and past editions of the IPC. A good overview of the deterministic
  track of the 2004 instance of the IPC can be found in
  \citep{Hoffmann-Edelkamp:05}.}  Using these planners---and an ad-hoc
Java implementation of GraphPlan \citep{Blum1997} which serves as a
baseline for certain experiments---we perform a series of tests on a
range of problem instances in our NLG domains.

Overall, our findings are mixed. On the one hand, we demonstrate that
some planners can readily handle the \emph{search} problems that arise
in our testing domains on realistic inputs, which is promising given
the challenging nature of these tasks (e.g., the sentence generation
task is NP-complete; see \citealt{KolStr02}). On the other hand, these
same planners spend tremendous amounts of time on \emph{preprocessing}
to analyse the problem domain in support of the search. On many of our
problem instances, the preprocessing time overshadows the search time.
For instance, FF spends 90\% of its runtime in the sentence generation
domain on grounding out actions and literals. As a consequence, we
conclude that the off-the-shelf planners we investigated are too slow
to be useful in real NLG applications.  We hope that our results will
spark some interest for more efficient implementations of planning
preprocessors, and offer our NLG domains and experiences as challenges
for the planning community.
